---
title: "Prediction of combination sensitivity scores (CSS) for drug combinations using drug target information and machine learning methods"
---

---
libraries needed for the execution of code
---

```{r} 
library( caret )
library( dplyr )
library( tidyr )
library( plyr )
library( tibble )
library( readr )
#library( glmnet )
```


---
function makes predictions and calculates R2, RMSE, MAE and correlation values
---

```{r}
get.prediction.summary<- function(  identifier, data, response, design, method, cv_p, numSeeds ) 
{ #input data should be a tibble/data.frame containing identifier (user information about experiment), data (list of features), response (CSS values), design ( specifies which features or feature combinations to use), method (machine learning method that was selected by user), cv_p (partition in cross-validation), numSeeds (number of seeds to be used)
  
  collect.stats<- function( featureMatrix, response, method, cv_p, seeds, designName ) 
  { #function to make predictions and calculate R2, RMSE, MAE and correlation values
    
    clStat <- list()
    resStat <- list()
    
    #calculate separately for every seed
    for ( i in 1:length( seeds ) )  {
      
      cat("Seed: ", i, '\n')
      
      #split the data
      set.seed( seeds[i] )
      intrain <- createDataPartition( y = response, p = cv_p, list = FALSE )
      
      training <- featureMatrix[intrain,]
      testing <- featureMatrix[-intrain,]
      
      
      set.seed( seeds[i] )
      trctrl <- trainControl( method = "cv", number = 10 )
      
      
      #select the method
      if ( method == 'glmnet' ) {
        
        # elasticnet
       
        set.seed( seeds[i] )
        myFit = train( training, as.vector( response[intrain] ), method = method,
                        trControl=trctrl )

       
      }
      
      if ( method == 'ranger' ) {
        
        set.seed( seeds[i] )
        myFit <- train( training,as.vector( response[intrain] ), method = method, trControl=trctrl )
        
      }
      
      
      if ( method == 'lm' ||  method == 'svmRadial' ) {
        
        set.seed( seeds[i] )
        myFit <- train( training, as.vector( response[intrain] ), method = method, trControl=trctrl )
        
      }
      
      #ridge is sensitive to zero variance columns
      if ( method == 'ridge' ) {
        
        zer_var <- nearZeroVar( training, saveMetrics = TRUE) 
        ind <- which( zer_var$zeroVar == 'TRUE' )
        
        if ( length( ind ) > 0 ) {
          
          cat("Removing zero variance columns: ", rownames( zer_var )[ind], '\n')
          
          set.seed( seeds[i] )
          new_training <- training[, -nearZeroVar( training ) ]
          testing <- testing[, -nearZeroVar( training ) ]
          myFit <- train( new_training, as.vector( response[intrain] ), method = method, trControl=trctrl )
          
          
        } else {
          
          set.seed( seeds[i] )
          myFit <- train( training, as.vector( response[intrain] ), method = method, trControl=trctrl )
          
        }
      }
      
      
      
      #build_predictions
      myPredict <- predict( myFit, newdata = testing )
      myCor <- cor( myPredict, as.vector( response[-intrain] ) )
      myMean <- mean( abs( myPredict - as.vector( response[-intrain] ) ) )
      myRMSE <- sqrt( mean( ( myPredict - as.vector( response[-intrain] ) )^2 ) )
      myR2 <- 1 - ( ( sum( ( myPredict - as.vector( response[-intrain] ) )^2  ) )/( sum( ( as.vector( response[-intrain] )- mean(as.vector( response[-intrain]) ) )^2  ) ) ) 
  
      clStat[[i]] <- data.frame( cor = myCor, mae = myMean, r2 = myR2 , rmse = myRMSE, seed = seeds[i], method = method, designNum = j, design = designName  ) 
      
      

  #MODEL  tmp_coeffs <- coef(myFit$finalModel, s = myFit$bestTune$lambda)
    
      #save coefficients
#      cl_coef[[i]] <- data.frame( Feature = as.character( tmp_coeffs@Dimnames[[1]][tmp_coeffs@i + 1]), coef = tmp_coeffs@x, seed = seeds[i], method = method, design_num = j, design = designName )
      #save importance
 #     cl_imp[[i]] <- data.frame( Feature = as.character( tmp_coeffs@Dimnames[[1]]), importance = c( 0, as.numeric( varImp(myFit)$importance$Overall  )  ), seed = seeds[i], method = method, design_num = j, design = designName )   
      
    }
    
    cat('\n')
    resStat <- ldply( clStat, data.frame )
    
    r <- list()
    
    r[[1]] <- resStat
    #r[[2]] <- res_coef
    #r[[3]] <- res_imp
    r
  }
  
  #print information about the user selections
  print( identifier )
  cat( "method: ", method, "\n" )
  cat( "# of seeds: ", numSeeds, "\n" )
  cat('\n')
  
  #save results for every design option
  resList <- list()  
 
  
  #loop over different design options  
  for ( j in 1:nrow( design ) )  
  {
    cat("Design: ", j , '\n')
    
    #find features or features combo to use
    ind <- which( design[j,] == 1 ) 
    
    designName <- paste( names( data )[ind], collapse = "+")
    ####
    cat("Design name: ", designName , '\n')
    ####
    designData <- do.call( cbind, data[ind] )
    
    indi <- which( colSums( designData ) == 0 | colSums( designData ) == ncol( designData ) )
   
    if ( length( indi ) >0 ) {
      designData <- designData[, -indi ] 
    }
    
   
    #specify the seeds
    seeds <- seq( 50, 10000, length.out = numSeeds )
    
    #collect statistics for every design specified by user
    f <- collect.stats( designData, response, method, cv_p, seeds, designName )
    
    resList[[j]] <- f[[1]]
    #res_coef[[j]] <- f[[2]]
    #res_imp[[j]] <- f[[3]]
  }
  
  res <- ldply( resList, data.frame )
  resList <- list( identifier = identifier, resStatisics = res, method = method )
}

```


---
function that plots R2, RMSE, MAE and correlation values for selected matrices of features
---

```{r}

plot.summary <- function( plotData ) 
{ #input data should be an output data from get.prediction.summary function
  
  icpl <-  ggplot2::ggplot( plotData, aes( seed, cor, colour = design ) ) + geom_point( size = 1.5, alpha = 0.7) + theme_bw() +
    labs(  x = "seeds ") + theme( plot.title = element_text( hjust = 0.5 ), text=element_text( family="Garamond", size=14 ), legend.text=element_text( size=4 ),axis.text.x = element_text(angle = 90, hjust = 1) , axis.title.y=element_blank() ) 
  
  icpl1 <- icpl + ggtitle( 'Correlation' ) 
  
  
  icpl <-  ggplot2::ggplot( plotData, aes( seed, mae, colour = design ) ) + geom_point( size = 1.5, alpha = 0.7) + theme_bw() +
    labs(  x = "seeds ") + theme( plot.title = element_text( hjust = 0.5 ), text=element_text( family="Garamond", size=14 ), legend.text=element_text( size=4 ),axis.text.x = element_text(angle = 90, hjust = 1) , axis.title.y=element_blank() ) + theme(legend.position="none")
  
  icpl2 <- icpl + ggtitle( 'MAE' )
  
  
  icpl <-  ggplot2::ggplot( plotData, aes( seed, rmse, colour = design ) ) + geom_point( size = 1.5, alpha = 0.7) + theme_bw() +
    labs(  x = "seeds ") + theme( plot.title = element_text( hjust = 0.5 ), text=element_text( family="Garamond", size=14 ), legend.text=element_text( size=4 ),axis.text.x = element_text(angle = 90, hjust = 1) , axis.title.y=element_blank() ) + theme(legend.position="none")
  
  icpl3 <- icpl + ggtitle( 'RMSE' )
  
  
  icpl <-  ggplot2::ggplot( plotData, aes( seed, r2, colour = design ) ) + geom_point( size = 1.5, alpha = 0.7) + theme_bw() +
    labs(  x = "seeds ") + theme( plot.title = element_text( hjust = 0.5 ), text=element_text( family="Garamond", size=14 ), legend.text=element_text( size=4 ),axis.text.x = element_text(angle = 90, hjust = 1) , axis.title.y=element_blank() ) + theme(legend.position="none")
  
  icpl4 <- icpl + ggtitle( 'R2' )
  
  grid.arrange(icpl3, icpl2, icpl4, icpl1, ncol = 4, widths = c( 1, 1, 1, 2 ) )
  
  
}


```


---
read CSS data and select a cell line for which predictions will be made
---

```{r} 

# download data
load( 'css_data.RData' ) 

cellLines <- unique( cssData$cellLine )
i = 2 #select the cell line number - OVCAR3 cell line 

cssData <- 
  cssData %>%
  filter( cellLine == cellLines[i] )

```


---
read fingerprint data and create a table with drug combination fingerprints
---

```{r} 

load( 'fingerprints.RData' ) 

a <- 
  cssData %>%
  left_join( fingerprints, by = c( "drug1" = "drug" ) ) %>%
  dplyr::select( - contains( 'drug' ), -cellLine, -CSS )
  
b <- 
  cssData %>%
  left_join( fingerprints, by = c( "drug2" = "drug" ) ) %>%
  dplyr::select(  - contains( 'drug' ), -cellLine, -CSS )
  
  
comboFingerprints <- as_tibble( a + b )

comboFingerprints <- 
  comboFingerprints%>%
  mutate_all( funs( ifelse( .== 2, 1, . ) ) )

```


---
read drug primary target data and create a table with drug combination primary targtes
---

```{r} 

load( 'primary_targets.RData' ) 

a <- 
  cssData %>%
  left_join( primaryTargets, by = c( "drug1" = "drug" ) ) %>%
  dplyr::select( - contains( 'drug' ), -cellLine, -CSS )
  
b <- 
  cssData %>%
  left_join( primaryTargets, by = c( "drug2" = "drug" ) ) %>%
  dplyr::select( - contains( 'drug' ), -cellLine, -CSS )
  
  
comboPrimaryTargets <- as_tibble( a + b )

comboPrimaryTargets <- 
  comboPrimaryTargets %>%
  mutate_all( funs( ifelse( .== 2, 1, . ) ) )

```


---
read drug primary and SEA target data and create a table with drug combination primary and SEA targtes
---

```{r} 

load( 'primary_plus_sea_targets.RData' ) 

a <- 
  cssData %>%
  left_join( primaryPlusSeaTargets, by = c( "drug1" = "drug" ) ) %>%
  dplyr::select( - contains( 'drug' ), -cellLine, -CSS )
  
b <- 
  cssData %>%
  left_join( primaryPlusSeaTargets, by = c( "drug2" = "drug" ) ) %>%
  dplyr::select( - contains( 'drug' ), -cellLine, -CSS )
  
  
comboPrimaryPlusSeaTargets <- as_tibble( a + b )

comboPrimaryPlusSeaTargets <- 
  comboPrimaryPlusSeaTargets%>%
  mutate_all( funs( ifelse( .== 2, 1, . ) ) )

```


---
call functions to make predictions, calculate R2, RMSE, MAE and correlation values and plot them
---

```{r} 

# put feature tables in a list object
listOfFeatures <- list( dataFingerprints = comboFingerprints , dataPrimaryTargets = comboPrimaryTargets, dataPrimaryPlusSeaTargets = comboPrimaryPlusSeaTargets )


#make a design matrix 

# 1 at first place stands for including dataFingerprints, 1 at second place stands for including dataPrimaryTargets, 1 at third place stands for including  dataPrimaryPlusSeaTargets
# you can combine s´different feature data in 1 design matrix, e.g.  c( 0,1,1) stand for design matrix that has dataFingerprints and dataPrimaryTargets
#designMatrix <- as.matrix( rbind( c( 1,0,0 ), c( 0,1,0 ), c( 0,0,1 ) ) ) # if you want to test every feature data separately

designMatrix <- as.matrix( rbind( c( 1,0,0 ), c( 0,1,0 ), c( 0,0,1 ), c( 1,1,0), c(1,0,1) ) )# if you want to test every feature data separately and test combinations of dataFingerprints + dataPrimaryTargets and dataFingerprints + dataPrimaryPlusSeaTargets


identifier <- list( cellLine =  cellLines[i], response = 'CSS', data = 'combination experiment' )# put any useful information that specifies the case

#make predictions and calculate R2, RMSE, MAE and correlation values
res <- get.prediction.summary( identifier, listOfFeatures, as.vector( cssData$CSS ), designMatrix, 'glmnet', 0.7, 5 ) # glmnet is selected machine learning method ( can be glmnet, ranger or svmRadial); 0.7 is partition in cross-validation, 5 is the number of seeds to be used

#plot the results 
plot.summary( res[[2]] )

```


